The Inconsistency in G\"odel's Ontological Argument: \\ A Success Story for AI in Metaphysics

### 2016-01-31 Papers/InProgress/IJCAI/submission.tex
Improve abstract and introduction

\typeout{An AI Success Story in Metaphysics: \\ the Case of the Inconsistency in G\"odel's Ontological Argument}

@@

\begin{abstract}
  This paper discusses the discovery of the inconsistency in G\"odel's
  ontological argument as a success story for artificial
  intelligence. Despite the popularity of the argument since the
  appearance of G\"odel's manuscript in 1970, the inconsistency of the
  axioms used in the argument remained unnoticed until 2013, when it
  was detected automatically by the higher-order theorem prover
  \textsc{Leo-II}. Understanding and verifying the refutation
  generated by the prover turned out to be a time-consuming task, and
  its completion, as reported here, required the reconstruction of the
  refutation in the Isabelle proof assistant. The development of an
  improved syntactical hiding for the embedding technique
  % , utilizing
  % Isabelle's binding notation mechanism, 
  allows the refutation to be
  presented in a human-friendly way, suitable for non-experts in the
  technicalities of higher-order theorem proving. This brings us a
  step closer to wider adoption of logic-based artificial intelligence
  tools by philosophers.

  In addition, the paper reports on further improvements regarding the
  full automation of the ontological argument in a higher-order modal
  logic \textbf{S5} with a universal accessibility relation.
\end{abstract}

@@

The in-depth computational analysis presented here substantially
extends previous computer-assisted studies of G\"odel's ontological
argument. Similarly to the related work \cite{J30,C40} the analysis has
been conducted with automated theorem provers for classical
higher-order logic (HOL; cf. \cite{andrewsSEP} and the references
therein), even though G\"odel's proof is actually formulated in
higher-order \emph{modal} logic (HOML; cf. \cite{homl} and the
references therein). To bridge between the two logics we utilise and
further improve the logic embedding approach \cite{J23,C40}, which has
already been employed successfully in preceding related work.

@@

Another novel contribution reported here (in
\S\ref{sec:improvedembedding}) is the implementation of an alternative
embedding for the slightly stronger modal logic \SFiveU, which is
based on a universal accessibility relation. Our experiments have
shown that the new embedding is more efficient, as the following two,
previously open problems can now be solved:
\begin{itemize}
\item Proving automatically the final theorem T3 \textit{Necessarily, there
    exists God} (in Scott's \shortcite{ScottNotes} consistent variant) directly from the axioms
  alone, without relying on the argument's intermediate argumentation
  steps (i.e., lemmata). 
% \marginpar{ToDo: 2004 is the year of Sobel's book, not of Scott's manuscript. We need to fix this as I fixed the date for Goedel's manuscript. But I don't know the correct date for Scott's manuscript.}
\item Verifying automatically (in Isabelle/HOL) the proof of the modal
  collapse \cite{Sobel}, which is one of the most strongly criticized
  `side-effects' of the argument's axioms.
\end{itemize}

>>>>

\typeout{The Inconsistency in G\"odel's Ontological Argument: \\ A Success Story for AI in Metaphysics}

@@

\begin{abstract}
  This paper discusses the discovery of the inconsistency in G\"odel's
  ontological argument as a success story for artificial
  intelligence. Despite the popularity of the argument since the
  appearance of G\"odel's manuscript in 1970, the inconsistency of the
  axioms used in the argument remained unnoticed until 2013, when it
  was detected automatically by the higher-order theorem prover
  \textsc{Leo-II}. 
  Understanding and verifying the refutation
  generated by the prover turned out to be a time-consuming task, and
  its completion, as reported here, required the reconstruction of the
  refutation in the Isabelle proof assistant and eventually led to a novel and more efficient way of automating higher-order modal logic \textbf{S5} with a universal accessibility relation. Furthermore, the development of an
  improved syntactical hiding for the embedding technique
  allows the refutation to be
  presented in a human-friendly way, suitable for non-experts in the
  technicalities of higher-order theorem proving. This brings us a
  step closer to wider adoption of logic-based artificial intelligence
  tools by philosophers.
\end{abstract}

@@

The in-depth analysis presented here substantially
extends previous computer-assisted studies of G\"odel's ontological
argument. Similarly to the related work \cite{J30,C40} the analysis has
been conducted with automated theorem provers for classical
higher-order logic (HOL; cf. \cite{andrewsSEP} and the references
therein), even though G\"odel's proof is actually formulated in
higher-order \emph{modal} logic (HOML; cf. \cite{homl} and the
references therein). To bridge between the two logics we utilise and
further improve the logic embedding approach \cite{J23,C40}, which has
already been employed successfully in preceding related work.

@@

Another novel contribution reported here (in
\S\ref{sec:improvedembedding}) is the implementation of an alternative
embedding for the slightly stronger modal logic \SFiveU, which is
based on a universal accessibility relation. Our experiments have
shown that the new embedding is more efficient, as the following two
previously open problems can now be solved:
\begin{itemize}
\item Automatically proving the final theorem T3 (\textit{Necessarily, there
  exists God}), directly from Scott's \shortcite{ScottNotes} (consistent) axioms
  alone, without relying on the argument's intermediate argumentation
  steps (i.e., lemmata).
\item Automatically verifying, in Isabelle/HOL, the proof of the modal
  collapse \cite{Sobel}, which is one of the most strongly criticized
  logical consequences of the argument's axioms.
\end{itemize}
